## Load libraries

```{r}
#| message: false
library(readr)
library(tidyverse)
```

## Set project path

```{r}
project_path <- "~/projects/group_11_project/"
```

## Load data

```{r}
expr_data <- read_csv(str_c(project_path, "data/_raw/Colorectal_Cancer_Gene_Expression_Data.csv"))
patient_data <- read_csv(str_c(project_path, "data/_raw/Colorectal_Cancer_Patient_Data.csv"))
```

## Combine the two datasets

### First make the datasets tidy

There are three interrelated rules which make a dataset tidy:

1.  Each variable must have its own column

2.  Each observation must have its own row

3.  Each value must have its own cell

Currently, in the expr_data, the columns represent observations and the rows variables. So it needs to be pivoted before the datasets are combined

```{r}
# Pivot expr_data
expr_data <- expr_data |>
  mutate(ID_GENE = ID_REF) |> 
  select(-ID_REF) |> 
  pivot_longer(-ID_GENE, names_to = "ID_REF", values_to = "Gene expression level") |> 
  pivot_wider(names_from = ID_GENE, values_from = "Gene expression level") 
```

### Now they can be combined:

```{r}
# Join the two datasets
data <- patient_data |>
  full_join(expr_data, by = "ID_REF")

write_tsv(data, 
          str_c(project_path, "data/01_dat_load.tsv"))
```
