## Load libraries

```{r}
#| message: false
library(readr)
library(tidyverse)
```

## Set project path and make data directory

```{r}
project_path <- "~/projects/group_11_project/"
data_dir <- "data/_raw/"

#Create folder for data:
if (!dir.exists(str_c(project_path,data_dir))) {
  dir.create(str_c(project_path,data_dir), recursive = TRUE)
}
```

## Load data

```{r}
<<<<<<< HEAD
expr_data <- read_csv("~/projects/project_group11/data/_raw/Colorectal_Cancer_Gene_Expression_Data.csv")
patient_data <- read_csv("~/projects/project_group11/data/_raw/Colorectal_Cancer_Patient_Data.csv")
=======
expr_file <- "Colorectal_Cancer_Gene_Expression_Data.csv"
patient_file <-"Colorectal_Cancer_Patient_Data.csv"
data_url <- "https://www.kaggle.com/api/v1/datasets/download/amandam1/colorectal-cancer-patients"
zip_file <- "raw_data.zip"
>>>>>>> aeac4d93490a6e225e53dfb9696841ac7727d803

#Download data and uzip
if( !file.exists(str_c(project_path, data_dir, expr_file)) ){
  download.file(
    url = data_url,
    destfile = str_c(project_path, zip_file)
    , mode = "wb")
  
  unzip(str_c(project_path, zip_file),
        exdir = str_c(project_path, data_dir))
}

#Remove the zip file
if (file.exists(str_c(project_path, zip_file))) {
  file.remove(str_c(project_path, zip_file))
}

#Load data into matrix
expr_data <- read_csv(str_c(project_path, data_dir, expr_file))
patient_data <- read_csv(str_c(project_path, data_dir, patient_file))
```

## Combine the two datasets

### First make the datasets tidy

There are three interrelated rules which make a dataset tidy:

1.  Each variable must have its own column

2.  Each observation must have its own row

3.  Each value must have its own cell

Currently, in the expr_data, the columns represent observations and the rows variables. So it needs to be pivoted before the datasets are combined

```{r}
# Pivot expr_data
expr_data <- expr_data |>
  mutate(ID_GENE = ID_REF) |> 
  select(-ID_REF) |> 
  pivot_longer(-ID_GENE, names_to = "ID_REF", values_to = "Gene expression level") |> 
  pivot_wider(names_from = ID_GENE, values_from = "Gene expression level") 
```

### Now they can be combined:

```{r}
# Join the two datasets
data <- patient_data |>
  full_join(expr_data, by = "ID_REF")

write_tsv(data, 
          str_c(project_path, "data/01_dat_load.tsv"))
```
