## Load libraries

```{r}
#| message: false
library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
```

## Set project path

```{r}
project_path <- "~/projects/group_11_project/"
```

## Loading the data:

```{r}
data <- read_tsv(str_c(project_path, "data/03_dat_aug.tsv"))
```

## Inspect the dataset:

```{r}
data
```

The final dataframe contains 62 observations(rows) and 1946 variables(columns). Each observation is a colorectal cancer sample. Of the 1946 variables, 1 is the sample ID, 10 are the sample metadata, and the remaining 1935 variables are different genes whose expression is measured.

The variables Age and Age_group contain the same information, just in a different format. The same goes for Dukes_stage and No_stage.

The variables DFS and DFS event both give information on Disease-Free Survival. DFS describes the length of time without the disease returning, and DFS event describes whether or not the disease returned.

```{r}
data |>
  count(DFS_event)
```

Out of the 62 patients, 25 showed disease free survival, while 37 showed a return of the disease.

## Describe the dataset

Patient Data - Numerical Variables

Age and DFS 'Simple statistics

```{r}
# Define a vector of numeric feature names
features_num <- c("Age", "DFS")
# Select numeric columns based on features_num
numeric_data <- data[ , features_num, drop = FALSE]

# Compute summary statistics for each numeric feature
summary_stats <- numeric_data |>
  summarise(across(everything(), list(
    min = min,
    max = max,
    mean = mean,
    sd = sd
  ), na.rm = TRUE))

#Format the summary statistics into a table
summary_table_long <- summary_stats %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Feature", "Statistic"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  arrange(Feature, Statistic)

#Transform back to wide format
summary_table_wide <- summary_table_long %>%
  pivot_wider(
    names_from = Statistic,
    values_from = Value
  )

# Print the wide table
print(summary_table_wide)
```

```{r}
# Select patient 
patient_data <- data[ 1:12]
patient_data <- patient_data|>
    mutate(Gender = case_when(Gender == 0 ~ "Male",
                            Gender ==  1 ~ "Female"))
patient_data
```

## Visualizing data

```{r}
# Check for NA values in the Gender column
sum(is.na(data$Gender))
```

```{r}
pl3 <- data |>
  ggplot(aes(x = factor(Gender), fill = factor(Gender))) + 
  geom_bar(stat = "count") +
  labs(
    title = "Gender Distribution", 
    x = "Gender", 
    y = "Count", 
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    axis.title.x = element_text(size = 14, face = "bold"),  # Modify x-axis title font size and make it bold
    axis.title.y = element_text(size = 14, face = "italic") # Modify y-axis title font size and make it italic
  ) +
  scale_x_discrete(labels = c("0" = "Male", "1" = "Female")) + 
  scale_fill_discrete(labels = c("0" = "Male", "1" = "Female")) +
  theme_classic()
pl3
```

From the Gender Distribution, we can studied on 62 patient samples, the number of males far exceeds the number of females, close to 3 times.

```{r}
pl<- data |>
  drop_na(Gender, Age_group)|>
  mutate(Gender = case_when(
    Gender == 0 ~ "Male",
    Gender == 1 ~ "Female"
  ))|>
  count(Gender, Age_group)|>
 ggplot(aes(x = Age_group,
             y = n,
             fill = Gender)) +
  labs(
    title = "The Gender Distribution Across Age Groups Chart", 
    x = "Gender", 
    y = "Count", 
    fill = "Gender"
  )+
  geom_col(position = position_dodge(
    preserve = "single"),
    colour = "black",
    alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major.x = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
       axis.text.x = element_text(angle = 45, hjust = 1), 
       axis.title.x = element_text(size = 10, face = "bold"),  # Modify x-axis title font size and make it bold
      axis.title.y = element_text(size = 10, face = "italic") # Modify y-axis title font size and make it italic
    )
pl
```

From the Patient Gender Distribution Across Age Groups Chart, we can studied on 62 patient samples and found that the patients were mainly over 40 years old .The most cases were in the 60-70 age group.

```{r}
pl1 <- data|>
  ggplot(aes(x = Dukes_stage, fill = Dukes_stage)) +
    geom_bar(stat = "count") +                       # Bar plot
    labs(title = "Dukes_Stage Distribution", x = "Dukes_stage", y = "Count") + # Labels
    theme_minimal() +                   # Clean theme
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x labels for better readability
    ) +
    theme_classic()# Print each plot
pl1
```

From the poit of Location Distribution, we can studied on 62 patient samples and found that the distribution of patients in Dukes_Stage is similar, with the largest number of patients in ‘C’ .

```{r}
pl2<- data|>
  ggplot(aes(x = Location,fill = Location)) +
    geom_bar(stat = "count") +                       # Bar plot
    labs(title = "Location Distribution", x = "Location", y = "Count") + # Labels
    theme_minimal() +                   # Clean theme
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x labels for better readability
    ) 
pl2
```

From the poit of Location Distribution, we can studied on 62 patient samples and found that the proportion of patients in colon is relatively less.

The relationship between age and DFS:

```{r}
ggplot(data, aes(x = Age, y = DFS)) +
  geom_point(color = "lightblue") +               # Scatter plot
  labs(
    title = "DFS vs Age",       # Plot title
    x = "Age",                  # X-axis label
    y = "DFS (months)"          # Y-axis label
  ) +
  theme_minimal() +             # Clean theme
  theme(
    panel.grid.major = element_line(color = "grey", size = 0.5),  # Grid lines
    panel.grid.minor = element_line(color = "grey", size = 0.25)
  )
```

The scatter plot illustrates the relationship between **age** and **DFS** .

-   **Trend**: There is no clear linear relationship, but younger patients (\<40 years) tend to have higher DFS values, suggesting better survival.

-   **Variation**: Patients aged between 60-70 show a wide range of DFS outcomes, from very low (\~10 months) to high (\~90 months), indicating greater variability in survival.

-   **Cluster**: Most data points are concentrated in the 50-70 age range with DFS between 20-60 months.
